---
title: Installing R, RStudio, and associated tools
---


## Installing or Updating R

### Install R

The R website is ... special. Since you probably don't know or care what a mirror is, here are links to the

- Go to the [R Project](https://cran.r-project.org/) website
- Choose the `Download and install R` your operation system:
    - Mac: Download `R-3.2.4.pkg`
    - Windows: First click on `base` and then on `Download R 3.2.4 for Windows`

### Upgrade R

If you already have R installed, to check that you have the newest version of R, run
```{r eval=FALSE}
sessionInfo()
```
which will produce output like,
```
R version 3.2.4 (2016-03-10)
Platform: x86_64-apple-darwin13.4.0 (64-bit)
Running under: OS X 10.11.3 (El Capitan)

locale:
[1] en_US.UTF-8/en_US.UTF-8/en_US.UTF-8/C/en_US.UTF-8/en_US.UTF-8

attached base packages:
[1] stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
[1] R6_2.1.2

loaded via a namespace (and not attached):
[1] rsconnect_0.4.2 tools_3.2.4     knitr_1.12.3
```
At the start of this course, March 29, 2016, the newest version of R is 3.2.4.
Check that your `R version` is that one.

### Update packages

Now, update all your packages
```{r eval=FALSE}
update.packages(ask = FALSE)
```
The option `ask = FALSE` answers "yes to all the questions" as to whether to answer


## Installing or Upgrading R RStudio

If you don't have RStudio installed, install it from [rstudio.com](https://www.rstudio.com/products/RStudio/#Desktop). You want to download the RStudio **Desktop Open Source Edition** (which is FREE).

If you already have RStudio installed, go to `Help > Check for Updates` in the menu and install a new version if the pop-up window indicates that there is a new version available.

## Installing Tools to Build R Packages from Scratch

Mature R packages are distributed over CRAN and installable in R using `install.packages` without any other assistance.
However, we may use some bleeding-edge R packages that require the packages to be built from source and installed using the [devtools](http://www.rstudio.com/products/rpackages/devtools/) package.
Install or

This requires some extra [programs](http://www.rstudio.com/products/rpackages/devtools/).

- Windows: Install the newest version of [Rtools](http://cran.r-project.org/bin/windows/Rtools/)
- Mac: Install Xcode from the Mac App Store.

To check that this is setup correctly, run the following
```{r eval=FALSE}
library("devtools")
has_devel()
```
If everything is correctly set up, it will return `TRUE`. If not, there will be an error.

## Installing LaTeX

LaTeX is needed to create PDFs from R Markdown files.
Install the following, depending on your operating system:

- Windows: [MikTex](http://miktex.org/download)
- OSX: [MacTex](https://tug.org/mactex/)

Note, these are very large files.
