---
title: "Loops and Conditionals"
---
## Learning Objectives

- Write and use `for` loops to apply the same code to multiple objects or elements
- Use `if`-`else` statements to conditionally execute code

## Required Packages and Datasets

Make sure you can load all these packages and dataset before starting the module
```{r, message = FALSE}
library(dplyr)
library(uwpols501)
data(turnout)
```

## For Loops

We use `for` loops to iterate through multiple elements of a list, variable, etc., and do **something** with those elements.

A `for` loop has the following skeleton:
```{r, eval = FALSE}
for (each_element in here) { # Try to use meaningful names for the
  # do something             #   "each_element" object
  print(each_element + 1)
}
```


A short/silly example:
```{r}
list_numbers <- sample(x = 1:100, size = 10, replace = FALSE) # sample() function
for (number in list_numbers) {
  print(paste0("2 times ", number, " is ", number * 2))
}
```

A real example:

Going back to the `turnout` dataset. Imagine that we want to estimate the same linear model for different groups of people based on their age: if they're in their 20s, 30s, etc. We can do that using a `for` loop and the `age_new` variable we just created. We want to estimate the following linear model:
```{r}
model_formula <- vote ~  educate + income
```

We first create an empty `results` dataset that we'll fill with the coefficients and SEs
of each model run.
```{r}
groups <- sort(unique(turnout$age_new))
results <- data.frame(age_new = groups,
                      edu_coef = rep(NA, length(groups)),
                      edu_se = rep(NA, length(groups)),
                      inc_coef = rep(NA, length(groups)),
                      inc_se = rep(NA, length(groups)))
```

Now we estimate the model for each age group and we store the coefficients and SEs for each covariate in the `results` dataset.
```{r}
for (i in 1:length(groups)) {
  gr <- groups[i]
  sub_data <- filter(turnout, age_new == gr)
  gr_model <- lm(model_formula, data = sub_data)
  results$edu_coef[i] <- coef(gr_model)[2]
  results$edu_se[i] <- sqrt(diag(vcov(gr_model)))[2]
  results$inc_coef[i] <- coef(gr_model)[3]
  results$inc_se[i] <- sqrt(diag(vcov(gr_model)))[3]
}
results
```

`r challenge_start()`
Take a look at the `results` dataset. Is there something wrong? Why? What happened? Talk to your classmates and try to figure it out (5 min.).
`r challenge_end()`

## Conditional Execution: if and else statements

Sometimes we only want to execute certain code if the data fulfills some conditions. To do that we use `if` and `else` statements.

How `if` and `else` statements look like:
```{r eval=FALSE}
if (this) {
  # do that
} else if (that) {
  # do something else
} else {
  #
}
```

**This** and **that** in the previous chunk of code are boolean tests: code that returns TRUE/FALSE when the computer executes it. Some examples of boolean tests:
```{r}
10 == 2
10 %in% c(2, 5, 13, 20, 10)
10 == 2 & 10 %in% c(2, 5, 13, 20, 10)
10 == 2 | 10 %in% c(2, 5, 13, 20, 10)
```

A short/silly example:
```{r}
some_numbers <- c(1, 4, 6, 10,12, 16, 45, 88, 102)
for (number in some_numbers) {
  if (number < 10) {
    print(paste0(number, " is smaller than 10"))
  } else if (number < 50) {
    print(paste0(number, " is smaller than 50 but greater or equal to 10"))
  } else {
    print(paste0(number, " is greater than 50"))
  }
}
```

Real example:

In the previous code where we estimated the same model for different groups of age we run into an issue with the last group: people in their 90s.
```{r}
table(turnout$age_new)
```

Since there are only 3 people who are in their 90s we run out of degrees of freedom. Remember that df = #observations - 1 - #variables_in_the_model. Let's adapt that code so that we only estimate the model in those cases we have enough degrees of freedom and prints a messages when that's not the case.
```{r}
results2 <- data.frame(age_new = groups,
                      edu_coef = rep(NA, length(groups)),
                      edu_se = rep(NA, length(groups)),
                      inc_coef = rep(NA, length(groups)),
                      inc_se = rep(NA, length(groups)))
for (i in 1:length(groups)) {
  gr <- groups[i]
  sub_data <- filter(turnout, age_new == gr)
  gr_model <- lm(model_formula, data = sub_data)
  if (gr_model$df > 1) {
    results2$edu_coef[i] <- coef(gr_model)[2]
    results2$edu_se[i] <- sqrt(diag(vcov(gr_model)))[2]
    results2$inc_coef[i] <- coef(gr_model)[3]
    results2$inc_se[i] <- sqrt(diag(vcov(gr_model)))[3]
  } else {
    print(paste0("Not enough obs. to estimate the model for gr == ", gr))
  }
}
results2
```