---
title: "Panel Data"
author: "Jeffrey Arnold"
date: "May 26, 2016"
---

## Clustered Standard Errors

Nunn and Wantchekon Data,

```{r}
library("dplyr")
library("broom")
library("plm")
library("lmtest")
library("multiwayvcov")
library("rio")
nunn <- import("data/Nunn_Wantchekon_AER_2011.dta")
```

OLS Model
```{r}
f <- trust_neighbors ~
        ln_export_pop +
        # controls
        # individual level
        age + age2 + male + urban_dum + education +
        occupation + religion + living_conditions +
        # district-level 
        district_ethnic_frac + frac_ethnicity_in_district +
        # country-level
        isocode
mod_nunn_ols <-
  lm(f, data = nunn)
tidy(mod_nunn_ols) %>%
  filter(term == "ln_export_pop")
```

Standard errors clustered by district
```{r}
coeftest(mod_nunn_ols, vcov. = function(x) cluster.vcov(x, ~ district)) %>%
  tidy() %>%
  filter(term == "ln_export_pop")
```
Standard errors clustered by ethnicity,
```{r}
coeftest(mod_nunn_ols, vcov. = function(x) cluster.vcov(x, ~ ethnicity)) %>%
  tidy() %>%
  filter(term == "ln_export_pop")
```

An alternative way to calculate cluster robust standard errors is the clustered-bootstrap.
Instead of resampling individual observations, we resample clusters.
See Assignment 3 
```{r}
ethnicities <- select(nunn, ethnicity) %>% unique() 
iter <- 1024 
bs_samples <- list() 
for (i in seq_len(iter)) { 
  # Resample ethnic groups 
  ethnic_sample <- sample_frac(ethnicities, 1, replace = TRUE) 
  # Merge with Nunn data to get a new dataset 
  newdata <- left_join(ethnic_sample, nunn, by = "ethnicity") 
  bs_samples[[i]] <- tidy(lm(f, data = newdata)) 
} 
bs_samples <- bind_rows(bs_samples) 
bs_samples %>% filter(term == "ln_export_pop") %>% 
  ungroup() %>%
  summarize(conf.low = quantile(estimate, 0.025), 
            conf.high = quantile(estimate, 0.975), 
            se = sd(estimate))
```





## Panel Data

Example from Michael Ross, "Is Democracy Good for the Poor?"

- Is there a relationship between democracy and, say, infant mortality? We could run a big cross-national regression, but would that be convincing?
- Perhaps democratic countries are different from non-democracies in ways that we can't measure---they are richer, provide benefits more efficiently, developed longer ago, or posses some cultural trait that tends to make their health outcomes better. 
- The idea in Ross's paper is to look at  countries over time to see if we can get a better estimate of the effect of democracy on infant mortality
- It turns out that under certain assumptions, we can allow for violations of zero conditional mean error if we have panel data (repeated observations over time)

```{r message=FALSE}
library("rio")
ross <- import("data/ross-democracy.dta")
```

In `ross`, the unit of observation is country (`id`), year (`year`).

The R package [plm](http://www.inside-r.org/packages/cran/plm) contains many functions for panel data.
```{r message=FALSE}
library("plm")
```

OLS ignoring fixed effects
```{r}
mod_lm <- lm(log(kidmort_unicef) ~ democracy + log(GDPcur), data = ross)
summary(mod_lm)
```
or using `plm`,
```{r}
mod_pooled <- plm(log(kidmort_unicef) ~ democracy + log(GDPcur),
              index = c("id", "year"),
              data = ross,
              model = "pooling")
summary(mod_pooled)
```

Fixed effects model using the "within" (de-meaned) estimator,
```{r}
mod_fe <- plm(log(kidmort_unicef) ~ democracy + log(GDPcur),
              index = c("id", "year"),
              data = ross,
              model = "within")
summary(mod_fe)
```

Fixed effects model using the least-squares dummy variable (LSDV) estimator,
```{r}
mod_fe_lsdv <- lm(log(kidmort_unicef) ~ democracy + log(GDPcur) + id,
              data = ross)
summary(mod_fe_lsdv)
```
Note that the coefficients on `democracy` are the same,
```{r}
coef(mod_fe)['democracy']
coef(mod_fe_lsdv)['democracy']
```


Two-way fixed effects model using the within estimator,
```{r}
mod_fe2 <- plm(log(kidmort_unicef) ~ democracy + log(GDPcur),
              index = c("id", "year"),
              data = ross,
              effect = "twoway",
              model = "within")
summary(mod_fe2)
```

Random effects model,
```{r}
mod_re <- plm(log(kidmort_unicef) ~ democracy + log(GDPcur),
              index = c("id", "year"),
              data = ross,
              model = "random")
summary(mod_re)
```

